Bow shocks, bow waves, and dust waves

### 2017-12-24 papers/Papers1234/sec-dust-bow-intro.tex
Nearly finished the i-front trapping (simpler now)

Then, defining a fiducial optical depth, \(\tau_* = \rho \kappa R_*\), and adding
in the stellar wind ram pressure\footnote{%
  We implicitly assume that the interaction of the stellar wind with
  the external stream can always be treated in the continuum limit.
  This will be true if either the collisional mean free path or the
  ion Larmor radius is much smaller than \(R_0\), which is almost
  always the case.} %
from equation~\eqref{eq:wind-efficiency}, we find that the general bow
radius can be written in terms of the fiducial radius as
\(R_0 = x R_*\), where \(x\) is the solution of
\begin{equation}
  \label{eq:rad-full-x}
  x^2 - \bigl(1 - e^{-2 \tau_* x} \bigr) - \eta = 0 \ .
\end{equation}
Since this is a transcendental equation, \(x\) must be found
numerically, but we can write explicit expressions for three limiting
cases:
\begin{equation}
  \label{eq:x-cases}
  x \approx
  \begin{cases}
    \text{if \(\tau_* \gg 1\):} & (1 + \eta)^{1/2}  \\
    \text{if \(\tau_*^2 \ll 1\):} & \tau_* + \bigl( \tau_*^2 + \eta \bigr)^{1/2} \approx
    \begin{cases}
      \text{if \(\tau_*^2 \gg \eta\):} & 2 \tau_*  \\
      \text{if \(\tau_*^2 \ll \eta\):} & \eta^{1/2} 
    \end{cases}
  \end{cases}
\end{equation}
The first case, \(x \approx (1 + \eta)^{1/2}\), corresponds to a
\textit{radiation bow shock}; the second case,
\(x \approx 2 \tau_* \), corresponds to a \textit{radiation bow wave}; and the
third case, \(x \approx \eta^{1/2}\), corresponds to a \textit{wind bow shock}.
The two bow shock cases are similar in that the external stream is
oblivious to the presence of the star until it suddenly hits the bow
shock shell, differing only in whether it is radiation or wind that is
providing the internal pressure.  In the intermediate bow wave case,
on the other hand, the external stream is gradually decelerated by
absorption of photons as it approaches the bow.\footnote{A shock does
  still form in this case, but shocked material constitutes only a
  fraction of the total column density of the shell, see
  \S~\ref{sec:shape-bow-wave}.}

@@

We wish to calculate whether the star is capable of photoionizing the
entire bow shock shell, or whether the ionization front will be
trapped within it.  The number of hydrogen recombinations\footnote{%
  The diffuse field is treated in the on-the-spot approximation,
  assuming all emitted Lyman continuum photons are immediately
  re-absorbed locally, so the case~B recombination co-efficient,
  \(\alphaB = \num{2.6e-13}\, (T/\SI{e4}{K})\, \si{cm^3.s^{-1}}\), is
  used.} %
per unit time per unit area in a fully ionized shell is
\begin{equation}
  \label{eq:shell-recombination-rate}
  \mathcal{R} = \alphaB n\shell^2 h\shell \ ,
\end{equation}
while the flux of hydrogen-ionizing photons
(\(h \nu > \SI{13.6}{eV}\)) incident on the inner edge of the shell is
\begin{equation}
  \label{eq:shell-ionizing-flux}
  \mathcal{F} = \frac{S} {4 \pi R_0^2} \ , 
\end{equation}
where \(S\) is the ionizing photon luminosity of the star.  Any shell
with \(\mathcal{R} > \mathcal{F}\) cannot be entirely photoionized by
the star, and so must have trapped the ionization front.  This simple
criterion is shown by the dark red line in
Figure~\ref{fig:zones-v-n-plane}.  If 
\(n / v_{10}^2 > \text{\numrange{1000}{5000}}\), depending
only weakly on the stellar parameters, then the outer parts of the
shocked shell are neutral, instead of ionized.

Assuming photoionization equilibrium, the
hydrogen photoabsorption optical depth of the shell is
\begin{equation}
  \label{eq:ion-tau-gas}
  \tau\gas = - \ln(1 - \mathcal{R} / \mathcal{F}) \ ,
\end{equation}
so long as \(\mathcal{R} < \mathcal{F}\).  The column density of the
shocked shell can be found, for example, from equations~(10) and~(12)
of \citet{Wilkin:1996a} in the limit \(v_\infty/V \to 0\) (Wilkin's parameter
\(\alpha\)) and \(\theta \to 0\).  This yields
\begin{equation}
  \label{eq:shocked-shell-column}
  n\shell h\shell = \tfrac34 n R_0 \ .
\end{equation}
Assuming strong cooling behind the shock,\footnote{%
  This is shown to be justified in \S~\ref{sec:radi-cool-lengths}.
} %
the shell density is \(n\shell = M_0^2 n\), where \(M_0 = v / \sound\)
is the isothermal Mach number of the external stream.  We will assume
a typical photoionized temperature of \SI{8000}{K}, so that
\(\sound \approx \SI{10}{km.s^{-1}}\) and \(M_0 = v_{10}\), yielding
\begin{equation}
  \label{eq:ion-tau-gas-expanded}
  \tau\gas = -\ln\bigl(1 -
  \num{8.42e-6}\, v_{10}^2 n^2 R_{0,\text{pc}}^3 S_{49}^{-1}\bigr) \ , 
\end{equation}
where 
\begin{align*}
  R_{0,\text{pc}} &= R_0 / \bigl( \SI{1}{pc} \bigr) \\
  S_{49} &= S / \bigl( \SI{e49}{s^{-1}} \bigr) \ . 
\end{align*}
Numerical values of \(S_{49}\) for our three example stars are given
in Table~\ref{tab:stars}.  The dust opacity is approximately constant
at FUV to EUV wavelengths, so the dust optical depth of the shocked
shell to ionizing photons follows from equations~\eqref{eq:tau-thin}
and~\eqref{eq:shocked-shell-column} as \(\tau\grain = \tfrac38 \tau\).

>>>>

Then, defining a fiducial optical depth,
\begin{equation}
  \label{eq:tau-star}
  \tau_* = \rho \kappa R_* \ ,
\end{equation}
and adding in the stellar wind ram pressure\footnote{%
  We implicitly assume that the interaction of the stellar wind with
  the external stream can always be treated in the continuum limit.
  This will be true if either the collisional mean free path or the
  ion Larmor radius is much smaller than \(R_0\), which is almost
  always the case.} %
from equation~\eqref{eq:wind-efficiency}, we find that the general bow
radius can be written in terms of the fiducial radius as
\(R_0 = x R_*\), where \(x\) is the solution of
\begin{equation}
  \label{eq:rad-full-x}
  x^2 - \bigl(1 - e^{-2 \tau_* x} \bigr) - \eta = 0 \ .
\end{equation}
Since this is a transcendental equation, \(x\) must be found
numerically, but we can write explicit expressions for three limiting
cases:
\begin{equation}
  \label{eq:x-cases}
  x \approx
  \begin{cases}
    \text{if \(\tau_* \gg 1\):} & (1 + \eta)^{1/2}  \\
    \text{if \(\tau_*^2 \ll 1\):} & \tau_* + \bigl( \tau_*^2 + \eta \bigr)^{1/2} \approx
    \begin{cases}
      \text{if \(\tau_*^2 \gg \eta\):} & 2 \tau_*  \\
      \text{if \(\tau_*^2 \ll \eta\):} & \eta^{1/2} 
    \end{cases}
  \end{cases}
\end{equation}
The first case, \(x \approx (1 + \eta)^{1/2}\), corresponds to a
\textit{radiation bow shock}; the second case,
\(x \approx 2 \tau_* \), corresponds to a \textit{radiation bow wave}; and the
third case, \(x \approx \eta^{1/2}\), corresponds to a \textit{wind bow shock}.
The two bow shock cases are similar in that the external stream is
oblivious to the presence of the star until it suddenly hits the bow
shock shell, differing only in whether it is radiation or wind that is
providing the internal pressure.  In the intermediate bow wave case,
on the other hand, the external stream is gradually decelerated by
absorption of photons as it approaches the bow.\footnote{A shock does
  still form in this case, but shocked material constitutes only a
  fraction of the total column density of the shell, see
  \S~\ref{sec:shape-bow-wave}.}

@@

We wish to calculate whether the star is capable of photoionizing the
entire bow shock shell, or whether the ionization front will be
trapped within it.  The number of hydrogen recombinations\footnote{%
  The diffuse field is treated in the on-the-spot approximation,
  assuming all emitted Lyman continuum photons are immediately
  re-absorbed locally, so the case~B recombination co-efficient,
  \(\alphaB = \num{2.6e-13}\, T_4^{-0.7}\, \si{cm^3.s^{-1}}\), is
  used, where \(T_4 = T/\SI{e4}{K}\).} %
per unit time per unit area in a fully ionized shell is
\begin{equation}
  \label{eq:shell-recombination-rate}
  \mathcal{R} = \alphaB n\shell^2 h\shell \ ,
\end{equation}
while the flux of hydrogen-ionizing photons
(\(h \nu > \SI{13.6}{eV}\)) incident on the inner edge of the shell is
\begin{equation}
  \label{eq:shell-ionizing-flux}
  \mathcal{F} = \frac{S} {4 \pi R_0^2} \ , 
\end{equation}
where \(S\) is the ionizing photon luminosity of the star.  Any shell
with \(\mathcal{R} > \mathcal{F}\) cannot be entirely photoionized by
the star, and so must have trapped the ionization front.  The column
density of the shocked shell can be found, for example, from
equations~(10) and~(12) of \citet{Wilkin:1996a} in the limit
\(v_\infty/V \to 0\) (Wilkin's parameter \(\alpha\)) and \(\theta \to 0\).  This yields
\begin{equation}
  \label{eq:shocked-shell-column}
  n\shell h\shell = \tfrac34 n R_0 \ .
\end{equation}
Assuming strong cooling behind the shock,\footnote{%
  This is shown to be justified in \S~\ref{sec:radi-cool-lengths}.
} %
the shell density is \(n\shell = \mathcal{M}_0^2 n\), where
\(\mathcal{M}_0 = v_\infty / \sound\) is the isothermal Mach number of the
external stream.\footnote{%
  \label{fn:temperature-dependence}
  The sound speed depends on the temperature and hydrogen and helium
  ionization fractions, \(y\) and \(y_{\text{He}}\) as
  \(\sound^2 = (1 + y + z_{\text{He}} y_{\text{He}}) (k T /
  \bar{m})\), where \(z_{\text{He}}\) is the helium nucleon abundance
  by number relative to hydrogen and
  \(k = \SI{1.3806503e-16}{erg.K^{-1}}\) is Boltzmann's constant.  We
  assume \(y = 1\), \(y_{\text{He}} = 0.5\), \(z_{\text{He}} = 0.09\),
  so that \(\sound = \num{11.4}\, T_4^{1/2}\, \si{km.s^{-1}}\). } %
Putting these together with equations~\eqref{eq:Rstar} and
~\eqref{eq:tau-star}, one finds that \(\mathcal{R} > \mathcal{F}\)
implies
\begin{equation}
  \label{eq:ifront-trap-x-cubed-taustar}
  x^3 \tau_* > \frac{4 S c \sound \bar{m}^2 \kappa}{3 \alpha L} \ .
\end{equation}
From equation~\eqref{eq:rad-full-x}, it can be seen that \(x\) depends
on the external stream parameters, \(n\), \(v_\infty\) only via
\(\tau_*\), and so equation~\eqref{eq:ifront-trap-x-cubed-taustar} is a
condition for \(\tau_*\).  In the radiation bow shock case,
\(x = (1 + \eta)^{1/2}\), and the condition can be written:
\begin{equation}
  \label{eq:ifront-trap-taustar-bow-shock}
  \tau_* > 145.0 \frac{S_{49} T_4^{1.7} \kappa_{600}}{L_4 (1 + \eta)^{3/2}} \ , 
\end{equation}
where
\begin{equation*}
  S_{49} = S / \bigl( \SI{e49}{s^{-1}} \bigr) \ .
\end{equation*}
Numerical values of \(S_{49}\) for our three example stars are given
in Table~\ref{tab:stars}.  In the radiation bow wave case,
\(x = 2\tau_*\), and the condition can be written:
\begin{equation}
  \label{eq:ifront-trap-taustar-bow-shock}
  \tau_* > \left(  18.1 \frac{S_{49} T_4^{1.7} \kappa_{600}}{L_4}\right)^{1/4} \ , 
\end{equation}

\(\tau_* \sim n^{1/2} / v_\infty\) 
This simple
criterion is shown by the dark red line in
Figure~\ref{fig:zones-v-n-plane}.  If 
\(n / v_{10}^2 > \text{\numrange{1000}{5000}}\), depending
only weakly on the stellar parameters, then the outer parts of the
shocked shell are neutral, instead of ionized.

Assuming photoionization equilibrium, the
hydrogen photoabsorption optical depth of the shell is
\begin{equation}
  \label{eq:ion-tau-gas}
  \tau\gas = - \ln(1 - \mathcal{R} / \mathcal{F}) \ ,
\end{equation}
so long as \(\mathcal{R} < \mathcal{F}\).

We will assume
a typical photoionized temperature of \SI{8000}{K}, so that
\(\sound \approx \SI{10}{km.s^{-1}}\) and \(M_0 = v_{10}\), yielding
\begin{equation}
  \label{eq:ion-tau-gas-expanded}
  \tau\gas = -\ln\bigl(1 -
  \num{8.42e-6}\, v_{10}^2 n^2 R_{0,\text{pc}}^3 S_{49}^{-1}\bigr) \ , 
\end{equation}
where 
\begin{align*}
  R_{0,\text{pc}} &= R_0 / \bigl( \SI{1}{pc} \bigr)
\end{align*}
The dust opacity is approximately constant
at FUV to EUV wavelengths, so the dust optical depth of the shocked
shell to ionizing photons follows from equations~\eqref{eq:tau-thin}
and~\eqref{eq:shocked-shell-column} as \(\tau\grain = \tfrac38 \tau\).
